---
title: "Time Series Project"
author: "Max Moro and Shane Weinstock"
output:
  html_document:
    df_print: paged
    toc: true
    toc_depth: 3
    toc_float: true
---

# Functions
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(tswge)
library(vars)
```

```{r}
ase = function(f,x){
  mean((f - tail(x,length(f)))^2)
}
```

# Data Loading and Transformation

```{r}
#Loading data
data = readRDS('TO_cc_data.rds') %>%
  filter(snap_fiscal_month_sort >= 201001
         ,snap_fiscal_month_sort <= 201909) %>%
  rename(period = snap_fiscal_month_sort)  %>%
  arrange(period) 
data$timeID = group_indices(data,period)
# Calculating Summaries
getSummary = function(x){x %>% 
    summarise(separationsCount = sum(cnt_non_RIF_separations,na.rm=T)
              ,attritionRate = sum(cnt_non_RIF_separations,na.rm=T)/sum(headcount,na.rm=T)
              ,headcount = sum(headcount,na.rm=T)
              ,ageMeanYrs = sum(tot_age,na.rm=T)/sum(headcount,na.rm=T)/12
              ,tenureMeanYrs = sum(tot_tenure,na.rm=T)/sum(headcount,na.rm=T)/12
              ,recognitionEvents = sum(spot_events+points_events+star_events,na.rm=T)
              ,recognitionEventsMean = recognitionEvents / headcount
              ,supervisorsCount = sum(cnt_spvs,na.rm=T)
              ,workersPerSupervisor = headcount / supervisorsCount
              ,lowPerfCount = sum(tot_ppa_low,na.rm=T)
              ,lowPerfRate = lowPerfCount / headcount
              ,highPerfCount = sum(tot_ppa_high,na.rm=T)
              ,highPerfRate = highPerfCount / headcount
              ,timeID = max(timeID)
    ) 
}
#calculating Totals
dataTot = data  %>%
  group_by(period) %>%
  getSummary() %>% ungroup()
#saving to X variable
x=dataTot$attritionRate
```

# 1 - Describe Data Set

The Data set represents the number of workers who voluntary left company A between 2010 and 2019. The data is real, but is anonymized to maintin the confidentiality of the company and its employees.

The business request is forecast of the attrition rate for the entire company for the next 12 months. 

The data set is composed of a single representation with multiple features that can be used to idenitify and build multi-variate models in order to forecast the attrition trend.

The data set has the following columns

Field|Description
----|-----
period| name of the period expressed as year and month
separationsCount| count of people who voluntary left the company
**attritionRate** | ratio between separationsCount and headcount (this is our target variable)
headcount| total number of workers active in the period
ageMeanYrs| average age of the active workers 
tenureMeanYrs | average tenure of the active workers 
recognitionEvents | count of all the monetary and non monetary recognitions provided to the workers during the period
recognitionEventsMean | average number of recognitions each worker received
supervisorsCount | count of all workers that manage other workers
workersPerSupervisor | average number of workers reporting to each supervisor
lowPerfCount | count of all workers that have a low perfomance annual rating
lowPerfRate | rate of low performers for the entire organization
highPerfCount | count of all workers that have an high perfomance annual rating
highPerfRate | rate of high performers for the entire organization
timeID | a unique ID per each period (1 for the first period up to 120 for the last period)

# 2 - The response from the Dataset

We selected the variable **attritionRate** as our response, it represents the ratio of people who have voluntary left the company each month.

# 3 - The Scenario

Attrition is an important factor for the company, as it operates in a very niche market, requiring workers with very advanced technical knowledge. Once an important worker has left the company, it takes many months to hire and train a replacement.The best solution to the problem is to predict the trend of attrition and attempt to counteract the correlated factors. The company can take action on these factors to reduce future attrition.

# 4 - MODELS

The models are trained through dataset with the exception of the most recent 12 months, we will use this portion to test each models performance. 

```{r define our test size}
# under variable TestSize, we have established we will test our data over the most recent 12 periods (measured in months).
TestSize = 12
```


## ARMA(3,2)  s=3

### a. ACF and Spectral analysis

```{r ACF and Spectral}
# initial plots to explore our current data
# all data is stored under variable "x"
p=tswge::plotts.sample.wge(x)
```

From the above graphs it can be observed that we have a wandering behavior in this realization. The ACF shows a quasi-cyclic behacior with a hint of a seasonality in the data.
The Parzen Window shows peaks at frequencies 0.17 and 0.34, which we can identify through their reciprocal as approximately 6 and 3 months.

Domain knowledge indicates that seasonality is expected, as people tend to voluntary leave the company during specific months (for example: when school ends or when financial results and bonuses are communicated). We are going to use a sesonality of **3 months** for this analysis as our domain knowledge finds it most appropriate.

### b. Check for seasonality

In order to eliminate any bias, we will establish through factor tables if the seasonal factor of s=3 or s=6 seasonality is in the data.

```{r}
tswge::factor.wge(phi=c(rep(0,5),1))
tswge::factor.wge(phi=c(rep(0,2),1))
e=tswge::est.ar.wge(x,p=10,type='burg')
```

we have a good evidence that the factors of **s=3** $(1-B^3)$ are present in the data. Frequencies from $(1-B^3)$ with value $0$ and $0.33$ are present in the data as 0,0 and 0.336, and the factors $(1-1B)$ and $(1+1B+1B^2)$ are present in the data as $(1-0.89B)$ and $(1-+0.8992B+0.754B^2)$. 


Before proceeding with the AIC5, we need to **remove the seasonality** ($s=3 months$)

```{r}
xtr=tswge::artrans.wge(x,phi.tr=c(rep(0,2),1))
```
**ACF and Frequency after transformation**

```{r}
p=tswge::plotts.sample.wge(xtr)
```

As the data has been transformed over the s=3 seasonality, we can proceed with the aic5 by using the **xtr** variable.

```{r AIC and BIC Testing}
aic5 = tswge::aic5.wge(xtr,p=0:5,q=0:3,type='aic')
bic5 = tswge::aic5.wge(xtr,p=0:5,q=0:3,type='bic')
```

**Top 5 models using AIC**
`r aic5`

**Top 5 models using BIC**
`r bic5`

**PACF**
```{r}
pacf(xtr)
```

**ACF**
```{r}
p=tswge::plotts.sample.wge(xtr)
```


The PACF shows an erratic patern up to AR(9), thus we can't be sure that there is a pure AR model that will fit our attrition trend. The  sample autocorrelation shows cyclical pattern extending above the value of 0.5, so we should expect to not have an MA only model but that we will need other components to model this particular trend.

The aic5.wge with AIC metric shows an ARMA(3,2) and ARMA(4,2) as our best models. The aic.wge with BURG metric shows an ARMA(3,2) and an AR(3) as good cadidates for the final model as well.

Therefore, we will proceed with exploring the following models:

- ARMA(3,2) with seasonality of 3 months
- ARMA(4,2) with seasonality of 3 months
- AR(3) with seasonality of 3 months



### c. Factored Form

```{r}
p=3;q=2;s=3
es=tswge::est.arma.wge(xtr,p=p,q=q)
```

**Phis**
```{r}
es$phi
if(p>1) tswge::factor.wge(phi=es$phi)
```
**Thetas**
```{r}
es$theta
if(q>1) tswge::factor.wge(phi=es$theta)
message('Variance: ',es$avar)
message('Mean: ',mean(x))
```

Upon conclusion of our exploraition, we've concluded that the following our is our final or best fit equation for this model:

$$(1-B^3)(1+0.3055B+0.2485B^2+0.2817B^3)(X_t-0.00734) = (1+0.9476B+0.9635B^2)a_t$$
$$\hat{\sigma}^2_a = 4.019*10^{-6}$$ 

### d. AIC

AIC for this model is  `r aic5[aic5[,1]==p & aic5[,2]==q,3]`


### e. ASE

```{r}
armaPred= tswge::fore.aruma.wge(x,phi = es$phi,theta = es$theta,s=s,n.ahead = TestSize,lastn = T)
armaASE = ase(armaPred$f,x)
```

The ASE of the model is `r format(round(armaASE,8),scientific=FALSE)`

## VAR

### a. Selecting multivariate data

As the attrition rate is a proportion of the headcount ($terminations / headcount$), we selected all the endogenous variables that are meaningful for the problem and are expressed as a rate of the headcount. This avoids using any variables that by design are correlated to headcount (example: number of supervisors) instead of the attrition rate. 

In addition, we noted that the recognition variable has data only for the past 2 years. We decided to remove this variable as it has too much missing data.

We specified a seasaonlity of 3 months as suggested by the ARMA model.

We are not specifying any exogenous variable as the company has no indication of future values for these variables.


```{r Selecting Multivariate Data}
X=dataTot %>% dplyr::select(attritionRate,ageMeanYrs, tenureMeanYrs,  workersPerSupervisor, lowPerfRate, highPerfRate)


XTrain = head(X,nrow(X)-TestSize)
XTest  = tail(X,TestSize) #12 months
```

### b. Checking Cross Correlations

**Attrition and age**
```{r fig.height=4, fig.width=8}
# cross correlation code chunk
par(mfrow=c(2,3),mar=c(3,2,3,2))
with(XTrain,{
  ccf(attritionRate,ageMeanYrs)
  ccf(attritionRate,tenureMeanYrs)
  ccf(attritionRate,workersPerSupervisor)
  ccf(attritionRate,lowPerfRate)
  ccf(attritionRate,highPerfRate)
})
```

The most significant correlations with the Attrition Rate are **Mean Age** with lags of 5 and -5, **tenure** has a significant correlation with lag 6, also **workers per supervisor** has a correlation at lag 5. **Low Performers Ratio** is correlated with a lag of -2.

We can deduce from the above cross correlations that the model may use lags up to 6.

```{r VAR Selection}
vars::VARselect(XTrain, lag.max = 7, type = 'trend',season = 3, exogen = NULL)
```

The model selected a lag order of 1 using a BIC measure ($HQ(n)$).


### c. Model Training

We noticed the model has a better performance and overall behavior when using higher number of lags, with the optimal at lag=6.

```{r}
varFit=VAR(XTrain,p=6,type='trend',season=3,exogen=NULL)
summary(varFit$varresult$attritionRate)
```

We noticed the model has a pvalue less than 0.05, hence we can reject the hypothesis the model is not significant

### d. AIC 

The AIC of the model is `r AIC(varFit)`

**AIC**
```{r AIC for VARfit}
AIC(varFit)
```

### e. Model Testing Plot

```{r VAR Prediction and Graph}
varPred = predict(varFit,n.ahead=TestSize)$fcst$attritionRate[,1]
varDataPred = cbind(dataTot,pred=NA)
varDataPred[(nrow(varDataPred)-TestSize+1):nrow(varDataPred),]$pred = varPred
ggplot(data=varDataPred,aes(x=timeID,y=attritionRate)) +
  geom_line() + 
  geom_line(aes(y=pred),color='red')

```
### f. ASE

```{r VAR ASE}
varASE = ase(varPred,tail(dataTot,TestSize)$attritionRate)
```

The ASE of the model is `r format(round(varASE,8),scientific=FALSE)`

## Neural Network

### a. Selecting multivariate data

We selected the same dataset used for the VAR section to establish our next set of variables.

```{r Establishing Time Series Train, Test, and All variables}
tsTrain = ts(XTrain,start=c(2010,1),frequency = 12)
tsTest = ts(XTest,start=c(2018,10),frequency = 12)
tsAll =   ts(rbind(XTrain,XTest),start=c(2010,1),frequency = 12)

```

### b. Model Training

We trained the Nerual Network Model by using 20 repetitions and using a difforder of 1 and 3. This reflected the trend and the seasonality we found in previous models. We set lag search parameter from 1 to 6 to incorporate the lag we found during the analysis for the VAR model, ensuring that we have all the components of our previous discoveries to yield the best model moving forward.

```{r}
set.seed(1701)
library(nnfor)
nnFit = nnfor::mlp(y=tsTrain[,'attritionRate'],reps=20
                   ,xreg = data.frame(tsAll[,colnames(tsAll) != 'attritionRate'])
                   ,difforder = c(1,3),hd.auto.type = 'cv',hd=NULL,lags=1:6)
nnFit
plot(nnFit)
```

The Neural Network Model used 5 regressors with a different selection of lags. 

### c. Model Testing Plot

```{r}
nnTest=forecast(nnFit,h=TestSize,xreg = data.frame(tsAll[,colnames(tsAll) != 'attritionRate']))
plot(XTest$attritionRate,type='l',ylim =c(0,0.011))
lines(seq(1,TestSize),nnTest$mean,col='red')
```

### d. ASE
```{r}
nnASE = mean((tsTest[,'attritionRate'] - nnTest$mean)^2)

```
The ASE of the model is `r format(round(varASE,8),scientific=FALSE)`


## Ensamble

The Neural Network model has great ASE performance, but it is not feasable for the business as we need to know the future values of the exogenous variables to calculate the attrition rate prediction. The business has no information about these future values. Furthermore, we encounter additional uncertainty as we try to predict these values as we have observed an increase the error rate and confidence interval upon modeling them. 

So the best models for the business are VAR and ARMA(3,2). 

We used a **Neural Network** algorithm to combine and improve the predictions of VAR and ARMA models. The model uses the predictions obtained from VAR and ARMA models as predictors, and the real attrition rate as target. This model can then be used to predict the future attrition values.

### a. Selecting Train and Test datasets

We used the first 2/3 of the train dataset to fit the VAR and ARIMA models, then the last 1/3 of the train dataset to fit the **ensamble model** based on the output of the first two models. We will then use the test dataset (12 months) to measure the performance of the ensemble model.

```{r}
#https://www.analyticsvidhya.com/blog/2018/06/comprehensive-guide-for-ensemble-models/
ensTrain = head(XTrain,nrow(XTrain)*2/3)
ensTrain2  = tail(XTrain,nrow(XTrain)*1/3)
```

### b1. Base Models Training

```{r}
## ARMA Training from the 2/3 of train dataset
ensARMA= tswge::fore.aruma.wge(ensTrain$attritionRate
                               ,phi = es$phi,theta = es$theta,s=s
                               ,n.ahead = nrow(ensTrain2),lastn = F)
ensARMAPred  = ensARMA$f

## VAR Training from  the 2/3 of train dataset
ensVAR= VAR(ensTrain,p=6,type='trend',season=3,exogen=NULL)
ensVARPred =  predict(ensVAR,n.ahead=nrow(ensTrain2))$fcst$attritionRate[,1]

## Plotting Predictions for the remaining 1/3 of the training dataset
plot(1:nrow(ensTrain2),ensTrain2$attritionRate, type='l',lwd=2)
lines(1:nrow(ensTrain2),ensARMAPred,col='red')
lines(1:nrow(ensTrain2),ensVARPred,col='blue')
```

We can see that the VAR prediction has a pattern similar to the prediction, while the ARIMA stays around the mean.

### b2. Ensemble Modeling
```{r}
## Training the ensambling (Neural Network Model)
set.seed(1701)
tsensTrain2 = tail(tsTrain,nrow(ensTrain2))
ensFit = nnfor::mlp(y=tsensTrain2[,'attritionRate'],reps=40
                    ,xreg = data.frame(ARMA = ts(ensARMAPred,start=min(time(tsensTrain2)),frequency=12)
                                       ,VAR = ts(ensVARPred,start=min(time(tsensTrain2)),frequency=12) )
                    ,difforder = NULL,hd.auto.type = 'cv',hd=NULL,lags=NULL)
ensFit
plot(ensFit)

```

The Neural Network Ensemble is using 2 regressors.

### c. Model Testing Plot

We are now proceeding to test the ensemble model for the Test dataset.

```{r}
#prediction from ARMA for the  test dataset
ensARMATestPred = tswge::fore.aruma.wge(XTrain$attritionRate
                               ,phi = es$phi,theta = es$theta,s=s
                               ,n.ahead =TestSize,lastn = F)$f

#prediction from   for the  test dataset
ensVARTest= VAR(XTrain,p=6,type='trend',season=3,exogen=NULL)
ensVARTestPred =  predict(ensVAR,n.ahead=nrow(XTest))$fcst$attritionRate[,1]

#prediction from Ensemble (combining the predictions from ARMA and VAR)
ensTest=forecast(ensFit,h=TestSize,xreg =data.frame(
  ARMA = c(ensARMAPred, ensARMATestPred)
  ,VAR =  c(ensVARPred, ensVARTestPred)))
  #ARMA =  ts(c(ensARMAPred, ensARMATestPred),start=min(time(tsensTrain2)),frequency=12)
  #,VAR =  ts(c(ensVARPred, ensVARTestPred),start=min(time(tsensTrain2)),frequency=12)))
plot(XTest$attritionRate,type='l',ylim =c(0,0.011))
lines(seq(1,TestSize),ensTest$mean,col='red')
```

we can see from the plot that the prediction looks very close to the real data.

### d. ASE
```{r}
ensASE = mean((tsTest[,'attritionRate'] - ensTest$mean)^2)
```

The ASE of the ensemble model is `r format(round(ensASE,8),scientific=FALSE)`

# 5. Comparing Models

Below are the ASE performance of the three models:

| Model | ASE |
|-------|-----|
| ARMA(3,2) | `r format(round(armaASE,8),scientific=FALSE)` |
| VAR | `r format(round(varASE,8),scientific=FALSE)` |
| Neural Network | `r format(round(nnASE,8),scientific=FALSE)` |
| Ensemble | `r format(round(ensASE,8),scientific=FALSE)` |

Based on the above ASE values, we can see that the Ensemble Model has the best performance.

# 6. Forecast

We've decided to use a forecast horizon of 12 months, as it is the most realistic intervention time for the company and the speed of the market in which the company is operating in. We are also going to use the **Ensemble Model** to predict future values. 

To run the ensamble model, we first need to create the predictions from the VAR and ARMA models, then we will use the fitted ensemble model to generate the future predictions.

NOTE: The package NNFOR doesn't provide confidence intervals for MLP function [link](https://r.789695.n4.nabble.com/Confidence-Intervals-with-mlp-forecasts-td4758966.html). We will plot, instead, the predictions from all the runs of the Neural Network model.

```{r}
horizon = 12
#prediction from ARMA
ensARMAFuturePred = tswge::fore.aruma.wge(X$attritionRate
                               ,phi = es$phi,theta = es$theta,s=s
                               ,n.ahead =horizon,lastn = F,plot=F)$f

#prediction from VAR
ensVARFuture = VAR(X,p=6,type='trend',season=3,exogen=NULL)
ensVARFuturePred =  predict(ensVARFuture,n.ahead=horizon)$fcst$attritionRate[,1]

#Ensamble model

##prediction from Ensemble
ensFuture=forecast(ensFit,h=TestSize,xreg =data.frame(
  ARMA =  c(ensARMAPred, ensARMAFuturePred)
  ,VAR =  c(ensVARPred, ensVARFuturePred))
  ,level=0.95)
#plot(ensFuture)
reps <- dim(ensFuture$all.mean)[2]
ts.plot(tsTrain[,'attritionRate'], ensFuture$all.mean, ensFuture$mean, col = c('black',rep("grey",reps), "blue"), lwd = c(1,rep(1, reps), 2))
#showLast = 50
#plot(tail(X$attritionRate,showLast),type='l',xlim=c(1,showLast+horizon))
#lines((showLast+1):(horizon + showLast),ensFuture$mean,col='red')
#lines((showLast+1):(horizon + showLast),ensFuture$low,col='red')
```

We can see the forecast follows the pattern and cycles of the previously recorded values, this is in line with the business expectations.